---
title: "Homework #2"
author: "Hugo Lhuillier"
date: "08/03/2018"
output: html_document
---

_Remember_: the cleaner and the more documented your code is, the more points you get.

## Exercice 1: playing with data 

The `sample.csv` file (in the `data` folder) contains the data of an experiment conducted on 100 patients. Patients belonged to one of the following groups: control, treatment 1 and treatment 2. For each patient, their level of aneurisms was recorded four times throughout the year.

With this data:

  1. [**easy**] load the `sample.csv` file (careful, it may contain columns name)
  1. [**easy**] currently, the `ID` variable is stored as factors. Convert this variable from factors to characters
  1. [**trivial**] as in the class, clean the `Gender` variable to get rid of `M` and `F`
  1. [**easy**] print summary statistics on the current data set 
  1. [**medium**] create a new variable, `Aneurisms_mean`, which corresponds to the mean aneurism across the four quarters, for each patient
  1. [**harder**] plot the distribution of `Aneurisms_mean` for both genders, and then for male and female separately


```{r, echo = TRUE}
# ANSWER HERE
d=read.csv(file="C:/Users/JIEJIE/Desktop/data/sample.csv",header=TRUE)
d$ID=as.character(d$ID)
typeof(d$ID)
d$Gender[d$Gender=="M"]="m"
d$Gender[d$Gender=="F"]="f"
table(d$Gender)
summary(d)
Aneurisms_mean=apply(d[,6:9],1,mean)
Aneurisms_mean
plot(Aneurisms_mean)
plot(Aneurisms_mean[d$Gender=="m"])
plot(Aneurisms_mean[d$Gender=="f"])
```

## Exercice 2: scaling to $(0,1)$

Write a function `rescale` that takes a vector as input and returns a corresponding vector of values scaled to lie in the range 0 to 1. 

```{r, echo = TRUE}
# ANSWER HERE
rescale=function(c){
  x = sqrt(sum(c*c))
  return(c/x)
}
#test#
rescale(c(1, 1, 1, 1))
```

## Exercice 3: testing

Design at least 3 tests -- using the `testthat` package -- that check whether your `scale_unit` function is working properly.

```{r, echo = TRUE}
# ANSWER HERE
test_that("testing the rescale function",{
  expect_true(all(rescale(c(1,2,3,100))<1))
})
```

# Exercice 4: scaling to $(a,b)$

Write a function `scale` that takes a vector with a range $(0,1)$ as input and returns a vector of values scaled to lie in the range `a` to `b`.

_Note_: if the original vector does not lie in the $(0,1)$ interval, warn the user that the computation is erroned.

```{r, echo = TRUE}
# ANSWER HERE
scaleab=function(c,a=0,b=1){
  if(all(c<1)){
    return(rescale(c)*(b-a)+a)
  }
  else print("the vector does not lie in the $(0,1)$")
}
scaleab(c(1,1,1,1),1,2)
scaleab(c(0.5,0.5,0.5,0.5),1,2)
scaleab(c(0.8,0.5,0.5,0.5),1,2)
scaleab(c(0.8,0.5,0.5,0.5))
```

# Exercice 5: composition

Write a function that scales an arbitrary vector `x` to a range `a` and `b`, where `a` and `b` are arguments of the function with default value `0` and `1`.

```{r, echo = TRUE}
# ANSWER HERE
nocheck_scaleab=function(x,a=0,b=1){
  return(rescale(x)*(b-a)+a)
}
scaleab(c(1,1,1,1),1,2)
scaleab(c(0.5,0.5,0.5,0.5),1,2)
scaleab(c(0.8,0.5,0.5,0.5),1,2)
scaleab(c(0.8,0.5,0.5,0.5))
```

# Exercice 6: visual checking

Using randown draws from a $(0,1)$ uniform distrbution, produce a plot that shows why using directly `scale` on a non-unit vector produces the wrong result. **Be creative**.

```{r, echo = TRUE} 
# ANSWER HERE
nonunit_scaleab=function(x,a=0,b=1){
  return(x*(b-a)+a)
}
u = nonunit_scaleab(2*runif(100),1,2)
plot(u)
'If the scale function applies on a non-unit vector, the value after scaling may not fall into the interval we want. In this case, suppose we have the vector only has one element. The value is randomly drawn from the uniform distribution between 0 and 2. We want to scale it into [1,2], but actually, there are points above 2, which are out of this interval and leads to the wrong result.'
```


