---
title: "Homework #2"
author: "Hugo Lhuillier"
date: "08/03/2018"
output: html_document
---

_Remember_: the cleaner and the more documented your code is, the more points you get.

## Exercice 1: playing with data 

The `sample.csv` file (in the `data` folder) contains the data of an experiment conducted on 100 patients. Patients belonged to one of the following groups: control, treatment 1 and treatment 2. For each patient, their level of aneurisms was recorded four times throughout the year.

With this data:

  1. [**easy**] load the `sample.csv` file (careful, it may contain columns name)
  1. [**easy**] currently, the `ID` variable is stored as factors. Convert this variable from factors to characters
  1. [**trivial**] as in the class, clean the `Gender` variable to get rid of `M` and `F`
  1. [**easy**] print summary statistics on the current data set 
  1. [**medium**] create a new variable, `Aneurisms_mean`, which corresponds to the mean aneurism across the four quarters, for each patient
  1. [**harder**] plot the distribution of `Aneurisms_mean` for both genders, and then for male and female separately


```{r, echo = TRUE}
# ANSWER HERE
```

## Exercice 2: scaling to $(0,1)$

Write a function `rescale` that takes a vector as input and returns a corresponding vector of values scaled to lie in the range 0 to 1. 

```{r, echo = TRUE}
# ANSWER HERE
```

## Exercice 3: testing

Design at least 3 tests -- using the `testthat` package -- that check whether your `scale_unit` function is working properly.

```{r, echo = TRUE}
# ANSWER HERE
```

# Exercice 4: scaling to $(a,b)$

Write a function `scale` that takes a vector with a range $(0,1)$ as input and returns a vector of values scaled to lie in the range `a` to `b`.

_Note_: if the original vector does not lie in the $(0,1)$ interval, warn the user that the computation is erroned.

```{r, echo = TRUE}
# ANSWER HERE
```

# Exercice 5: composition

Write a function that scales an arbitrary vector `x` to a range `a` and `b`, where `a` and `b` are arguments of the function with default value `0` and `1`.

```{r, echo = TRUE}
# ANSWER HERE
```

# Exercice 6: visual checking

Using randown draws from a $(0,1)$ uniform distrbution, produce a plot that shows why using directly `scale` on a non-unit vector produces the wrong result. **Be creative**.

```{r, echo = TRUE} 
# ANSWER HERE
```


