---
title: "Homework #3"
author: "Hugo Lhuillier"
date: "08/03/2018"
output: html_document
---

_Remember_: the cleaner and the more documented your code is, the more points you get.

## Exercice 1: factorial

Recall that the factorial of $n$, written $n!$, is defined as $n! = n \times (n - 1) \times \dots \times 2 \times 1$.

The `R` function `factorial(n)` compute the factorial of `n`. Using a `for` loop, write your own version of this function. 

_Hint_: similar to the `sum` function we wrote in class.

```{r, echo = TRUE}
# ANSWER HERE
```

## Exercice 2: binomial distribution 

The binomial random variable `Y \sim Bin(n,p)` repesents the number of successes in `n` binary trials, where the probability of success of each trial is the probability `p`.

Using only the `runif()` command - draw a number from the $(0,1)$ uniform distribution, write a function such `own_binomial` such that `own_binomial(n,p)` generates one draw of `Y`.

_Hint 1_: if $U \sim (0,1)$, and $p \in (0,1)$, then the expression `U < p` evalutes to `true` with probability `p`.

_Hint 2_: you will need to use both a `for` loop and a `if` statement.

```{r, echo = TRUE}
# ANSWER HERE
```

## Exercice 3: binomial distribution (test)

The `R` function `rbinom(n,size,p)` generates `n` draws from a binomial distribution $Bin(size,p)$ _[i know, i was confused myself]_. 

Using this function, and your function defined in exerice 2, test if your function is asymptotically correct.

_Hint_: you can look [here](https://www.r-bloggers.com/overlapping-histogram-in-r/) for some tricks on how to generate overlapping histrograms.

```{r, echo = TRUE}
# ANSWER HERE
```

## Exercice 4: random device 

Write a program that compute one realization of the following random device:

1. flip an unbiased coin 30 times (_hint_: once again, using `runif`)
1. whenever 3 consecutive heads occur, add one dollar to your wealth (once again, very similar to how we computed the sum in class)
1. whenever 3 consecutive tails occur, remove one dollar from your wealth

```{r, echo = TRUE} 
# ANSWER HERE
```
