---
title: "Homework #3"
author: "Julie Lenoir"
date: "17/03/2018"
output: html_document
---

## Exercice 1: factorial

Recall that the factorial of $n$, written $n!$, is defined as $n! = n \times (n - 1) \times \dots \times 2 \times 1$.

The `R` function `factorial(n)` compute the factorial of `n`. Using a `for` loop, write your own version of this function. 

```{r, echo = TRUE}
factorial = function(n){
  a = 1
  for(t in 2:n){
    #we set the loop so that at each step, we multiply a by     a + 1
    a <- a * t
  }
return(a)
}
factorial(3)
```

## Exercice 2: binomial distribution 

The binomial random variable `Y \sim Bin(n,p)` repesents the number of successes in `n` binary trials, where the probability of success of each trial is the probability `p`.

Using only the `runif()` command - draw a number from the $(0,1)$ uniform distribution, write a function such `own_binomial` such that `own_binomial(n,p)` generates one draw of `Y`.

```{r, echo = TRUE}

own_binomial = function(n, p){
  if(p<0 | p>1){
    warning("p is representating the probability of success, it needs to lie in [0, 1]")
  }
else if(0<=p && p<=1){
  #define a empty vector of results
  results = c(rep(NA, n))
    for(t in 1:n){
      #add 1 as an element of our results vector with probability p
      results[t] <- as.integer(runif(1)<p)
    }
  return(results)
  }
}


own_binomial(10, 0.5)
```

## Exercice 3: binomial distribution (test)

The `R` function `rbinom(n,size,p)` generates `n` draws from a binomial distribution $Bin(size,p)$ _[i know, i was confused myself]_. 

Using this function, and your function defined in exerice 2, test if your function is asymptotically correct.

```{r, echo = TRUE}

library(ggplot2)

#Let's constitute our database to compare the results of the two functions: own_binomial and rbinom

freq = c()
prob = c()
for(p in c(0.1, 0.3, 0.5, 0.7, 0.9)){
  own = unname(table(own_binomial(1000,p)) )
  R = unname(table(rbinom(1000,1, p)) )
  freq = rbind(freq, t(t(own)), t(t(R))  )
  prob = rbind(prob, matrix(rep(p), 4, 1) )       
}
type = matrix(rep(c("f", "t", "F", "T")), 20, 1)
data = data.frame(type = type, freq=freq, proba=prob)

legend = c("Failures from\nown_function", "Failures\nfrom R", "Successes from\nown_function", "Successes\nfrom R")

ggplot( data, aes(x=type, y=freq) ) + 
         facet_wrap(~proba) +
         geom_bar(aes(fill = type), 
                  stat="identity", 
                  position = "dodge") +
         labs(title="Comparing our own binomial function and the R one", y="Number of Success and Failures", x="") + scale_fill_discrete(name = "Legend", labels = legend) + theme(legend.position="bottom")

```

## Exercice 4: random device 

Write a program that compute one realization of the following random device:

1. flip an unbiased coin 30 times (_hint_: once again, using `runif`)
1. whenever 3 consecutive heads occur, add one dollar to your wealth (once again, very similar to how we computed the sum in class)
1. whenever 3 consecutive tails occur, remove one dollar from your wealth

```{r, echo = TRUE} 

gain = function(n){

#we set the zero variables and empty vectors we want to build on
results = c()
wealth = 0
t = 1
i = 2

#we loop over to get results :

for(t in 1:30){
  if(runif(1)<0.5){
    results = cbind(results, "Tail")
  }
  else{results = cbind(results, "Heads")}
  t <-  t+1
}

#we now compute final wealth
#if R finds 3 "Heads"/"Tail" in a row, he will next seek for a triple at i = i+3 (otherwise if we get 4 heads/tails, we would gain/loose 2 instead of 1)
#if R doesn't find 3 "Heads"/"Tail" in a row, he will next seek for a triple at the next element of Results

while(i <= 29){
  if(all(results[(i-1):(i+1)] == c(rep("Heads",3)))){
    wealth <- wealth + 1
    i = i+3

  }
  else if(all(results[(i-1):(i+1)] == c(rep("Tail", 3)))){
    wealth <- wealth - 1
    i = i+3
  }
  else{i <- i+1}
}
wealth = floor(wealth)
print(as.vector(results))
return(wealth)
}
gain(30)



```
