---
title: "Homework #5"
author: "Julie Lenoir"
date: "02/04/2018"
output: html_document
---

```{r, echo = FALSE, warning=FALSE,message=FALSE}
require(tidyverse)
require(Lahman)
require(zoo)
require(knitr)
theme_set(theme_bw())
```


# Exercice: American vs. National League

**Q1**: the American league was created in 1901. Subset the dataset so that it contains only observations when both leagues are present.

```{r, echo = T}

# Importing DataBase
team <- as.tibble(Lahman::Teams)

# Subsetting the data to get only observations for American league and National one

team <- filter(team, yearID>=1901, lgID != "FL")

```

**Q2**: compute the average runs per game for each team, each year

```{r, echo = TRUE}

# Use a pipe to compute the mean (use kable to display the table nicely)

kable(head(team %>%                       #selecting the databse
          group_by(teamID, yearID) %>%    #order the columns
              summarize(av_run = R/G)     #compute the mean
  )
)

```

**Q3**: create a new data frame that contains observations at the league level. In particular, we want the number of games and the number of runs, per league (`AL` or `NL`) and per year. Once this is done, compute the average runs per game (again, for each league, each year) 

```{r}

# Doing both at the same time : creating the dataframe + obtaining average runs per games (useful for next questions)

Nteam = team %>% 
  group_by(lgID, yearID) %>% 
    summarize(tot_run = sum(R), tot_game = sum(G), av_run = tot_run / tot_game)


kable(head(Nteam))


```

**Q5**: plot the time series of average runs per game, for each league.

```{r}

# Defining the plot G (in order to use it again for next question)

G = ggplot() +
  geom_point(data = Nteam, aes(yearID, av_run, colour=lgID, shape=lgID), stat="identity") + 
  labs(title="Time series of average runs per game, by league",
       x = "Time (years)", 
       y = "Average number of runs per game",
       colour ="League",
       shape = "League")
G
```

As you can see, the time series are noisy. To get a better picture, we are going to compute the 10-year moving average. To do that, we will use the function `rollmean` from the `zoo` package (if you don't know how to use it, `?roomean` should help).

**Q6**: compute, for each league, the 10-year moving average of the runs per game variable. Once this is done, plot together the time series of the original average runs per game (same as before), and the the 10 year moving average of the runs per game.

```{r}

# Computing the 10-year moving averages by league

mov_av_AL = rollmean(filter(Nteam, lgID=="AL")$av_run, 10)
mov_av_NL = rollmean(filter(Nteam, lgID=="NL")$av_run, 10)

# Creating a dataframe to use ggplot easily

mov_av = data.frame(
  year = rep(1905:2011,2),
  league = c(rep("AL", 107), rep("NL", 107)),
  mov_av = c(mov_av_AL, mov_av_NL))

# Creating the plot (G is the plot we created in last question)

G +
  geom_line(data = mov_av, mapping = aes(year, mov_av, colour=league)) +
  labs(subtitle = "Smoothed with moving average", 
       legend.title="League")
```

**Q7**: create a new dataset that computes the difference between the average runs per game (both for the original data and the 10-year moving average) of the American league and the National league, for each year.

```{r, message=FALSE}

# Computing averages and moving averages per league using our two databases

AL_av = Nteam %>%
  group_by(lgID, yearID) %>%
    filter(lgID == "AL") %>%
      select(av_run)

NL_av = Nteam %>%
  group_by(lgID, yearID) %>%
    filter(lgID == "NL") %>%
      select(av_run)

AL_mav = mov_av %>%
    group_by(league, year) %>%
      filter(league == "AL") %>%
        select(mov_av)
  
NL_mav = mov_av %>%
    group_by(league, year) %>%
      filter(league == "NL") %>%
        select(mov_av)

# Creating a global DataBase that contains all variables of interest

diff = data.frame(
  year = rep(1901:2016),
  av_AL = AL_av$av_run,
  av_NL = NL_av$av_run,
  mav_AL = c(rep(NA, 5), AL_mav$mov_av, rep(NA, 4)),
  mav_NL = c(rep(NA, 5), NL_mav$mov_av, rep(NA, 4))
)

# Computing the differences : NL - AL

diff = mutate(diff, 
              dav = av_NL - av_AL, 
              dmav = mav_NL - mav_AL)

kable(head(diff))
```

**Q8**: you now have all the data needed to reproduce the original plot. Specifically, plot the time series of the difference of the 10-year moving average, and superpose to it the actual difference (i.e. not the 10-year moving average).

```{r}

# Creating the plot

ggplot(diff) + 
  geom_bar(aes(year, dav), stat="identity", alpha=0.2, width=0.4) + 
  geom_line(aes(year, dmav), color = "darkred") + 
  labs(title = "Run scoring trend: NL difference from AL",
       subtitle = "1901-2016",
       x = "Year",
       y = "Runs per game")

```