---
title: "Introduction to Programming"
subtitle: "<h3> *R: Data types* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---
```{r setup, cache = F, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

# Object-oriented programming language in a nutshell

## Looks complicated, but `err`

- OOP: programming methodology based on objects, instead of just functions
- Oject: pretty much anything and everything
- Classes: a category of objects; defines the common properties of the objects that belong to this class 
- An object: an instance of a class $\Rightarrow$ different objects from an identical class have similar structure, but can be assigned individual characteristics 
- Can create relationships between objects (e.g. can inherit properties from one another etc.)
- Can create functions that apply only on some objects

## OOP: an example 

- Let's `Human` be a class, with three characteristics 
```{YAML, eval = FALSE}
Human: 
  - Name        # will have to be a string
  - Age         # will have to be a numeric
  - Address     # would be nice to have a common structure for addresses
```
- Let's `Address` be another class 
```{YAML, eval = FALSE}
Address: 
  - Number       # will have to be a numeric
  - Street       # will have to be a string
  - City         # will have to be a string
```
- Then, create my own address and human (two objects)
```{julia, eval = FALSE}
fredo-address = Address("27", "rue Saint-Guillaume", "Paris")
fredo-mion    = Human("Frédéric Mion", 49, fredo-address)
```
- And could also define functions
```{julia, eval = FALSE}
function where-i-live(address)
  "address.numer, address.street, address.city"

function presentation(human)
  "Hello, I am human.name. I am human.age, and I live where-i-live(human.address)"
```

<!-- 
  1. name, age and address are all attributes of the class Human (and the corresponding objects created from this class) 
  1. we have nested the Address class inside the Human class
  1. fredo-address and fredo-mion are repsecitvely two objects, instances of Human and Address. These are created by "constructors"
  1. the function where-i-live applies only on Address objects, and presentation on Humans. if were to put human inside where-i-live, would not work (would not make any sense right?)
-->

## Useful? 

- OOP languages are more flexible and allows you to write code that is extremely reusable
- **Example**: two functions can have the same names, but apply on different objects, and therefore produce different output
```{r, fig.align='center', echo = TRUE}
x <- 1:10
plot(x)
```

## Useful? 

- **Example**: two functions can have the same names, but apply on different objects, and therefore produce different output
```{r, fig.align='center', echo = TRUE, warning = FALSE}
x   <- runif(100); y <- 3 * x + rnorm(100)
reg <- lm(y ~ x + y)
par(mfrow=c(2,2))
plot(reg)
```
```{r, include = FALSE} 
par(mfrow=c(1,1))
```

# R and OOP

## `R`'s raw classes 

- `R` has 6 raw classes

    - `character`: `"a"` and `"abc"`
    - `numeric` : `1` and `1.5`
    - `integer`: `1L`
    - `logical`: `TRUE`
    - `complex`: `1.4i`
    
- In `R`, **everything** is an object 
```{r, echo = TRUE}
x <- 2.0
```
- `x`: instance of the numeric class, with zero attribute
 
## `R` raw classes 

- `R` has several functions to examine what objects we are dealing with

    - `class()` and `typeof()`: what kind of object is it
    - `length()`: how long is it
    - `attributes()`: what are its attributes
    
```{r, echo = TRUE}
x <- 1:10
typeof(x)
length(x)
attributes(x)
```

# `R`'s data type 

## `R`'s special values 

- `R` contains certain objects that makes life of data scientist easier

    - `NA`: missing value - can be incorporated in any vector, regardless of the data the latter contains 
    - `Inf`: infinity
    - `NaN`: not a number, i.e. an undefined value 
    
```{r, echo = TRUE}
x <- c(1, NA)
```
```{r, echo = TRUE}
1 / 0
0 / 0
```

## Data structures 

- Data structures are other classes that groups together `R`'s raw classes

    - atomic vector 
    - list 
    - matrix 
    - data frame
    - factors
    
## Atomic vectors 

- Vectors are the most basic structures in `R`
- A vector is a collection of elements, that can be `character`, `logical`, `integer` etc.
```{r, echo = TRUE}
vector()                        # create an empty logical (by default) vector
logical(0)                      # id. to the above constructor
vector("numeric", length = 5)   # create a numeric vector of length 5 
numeric(5)                      # id. to the above constructor
```
<!-- remember the Human and Address constructors above? Exactly id. here-->
- Alternatively, can directly create a vector with its content, and `R` will guess the appropriate mode of storage 
```{r, echo = TRUE}
x <- c(1.0, 2.0, 3.0)
x <- c(x, 4.0)          # and can combine vectors via the function c() 
```

## Note: `R`'s coercion 

- Remember: `R` is an interpreted languages
```{r, echo = TRUE}
x <- c(1.0, 2.0)
x <- c("a", "b")
```
<!-- therefore do not have to precise what types of data the variable x will store, and R can convert the variable R to a different mode of storage -->
- "Coercion": conversion between the different modes of storage
- Above = implicit coercion, and below explicit coercion 
```{r, echo = TRUE}
x <- 1:10
typeof(x)
y <- as.numeric(x)
typeof(y)
z <- as.character(y)
typeof(z)
```

## Matrix 

- Matrix: extension of vectors $\Leftrightarrow$ an atomic vector with dimensions
- Created via the constructor `matrix`
- Matrix are filled column-wise by default, but can change this by setting `byrow = TRUE`
```{r, echo = TRUE}
x <- matrix(1:10, ncol = 5, nrow = 2); print(x)
x <- matrix(1:10, ncol = 5, nrow = 2, byrow = TRUE); print(x)
dim(x)                               
```
-  Note: if apply `dim()` to an object that isn't a matrix, returns `NULL`

## Matrix 

- Can also create matrices by combining rows and/or columns via `cbind` and `rbind`
```{r, echo = TRUE}
x <- 1:3; y <- 10:12 
cbind(x,y)
rbind(x,y)
```
- **Question**: what happens if one vector is longer than another?

## List 

- Lists (aka generic vector)
- Unlike atomic vector, not restricted to a single to a single data type
- As you guessed, create list via the corresponding constructor `list` 
```{r, echo = TRUE}
x <- list(1, "a", TRUE, c(1, 2, 3)); print(x)
```
- The elements in a list are retrieved by using doubel square brackets
```{r, echo = TRUE}
x[[4]]
```

## List 

- Elements of list can be named 
```{r, echo = TRUE}
x <- list(a = "Fredo", b = "Mion", c = matrix(1:16, nrow = 4, ncol = 4)); print(x)
names(x)
attributes(x) 
```
<!-- here an object with attributes! -->
- When named, `x[[3]] == x$c`

<!-- Lists can be extremely useful inside functions. Because the functions in R are able to return only a single object, you can “staple” together lots of different kinds of results into a single object that a function can return. -->

## List: exercice 

> - What's the class of `x[[3]]`?
>- A matrix! 
>- What's the class of `x <- c("a", 1.0)`?
>- A vector of character! `1.0` $\Rightarrow$ `"1.0"`. Atomic vectors can only contains one type of class; `R`picks the class that can accomodate both types
>- What's the class of `x <- c(TRUE, 1.0)`?
>- A vector of numeric, with `TRUE` $\Rightarrow$ `1.0`
>- let `x = list(TRUE, 1.0)`. What's the type of `x[[1]]`?
>- A logical
>- **Careful**: sometimes, coercion is not feasible

## Data Frame

- How do we create data frame? `data.frame` clearly
```{r, echo = TRUE} 
x <- data.frame(id = 1:10, income = runif(10))
```
- Note: data frames are special lists $\Rightarrow$ same attributes and functions + new ones (e.g. `str`)
```{r, echo = TRUE}
attributes(x)           # names' one of its attributes, along others
x$id                    # = x[["id"]]
str(x)                  # able to store multiple data class 
```
<!-- str() is an instance of another function that is defined only on some object -->

## Summary of data structures

| Dimension | Homogenous class |    Heterogeneous class  |
|:---------:|:----------------:|:-----------------------:|
|     1d    |   Atomic vector  | List (= generic vector) |
|     2d    |      Matrix      |      Data frame         |

## Factors

- Factors = categorical variables in econometrics terms
<!-- remember that factors are a certain data structure, not a raw class!-->
- Factors are **stored as integers**, and have labels associated with these unique integers 
- Factors can only contain pre-defined values = _levels_ 

## Factors 

- Below is a factor with 3 levels
```{r, echo = TRUE}
x <- factor(c("low","high","medium","medium","high","low"))
```
- Levels are created by default in alphabetical order: 

    1. = "high"
    2. = "low"
    3. = "medium"

```{r, echo = TRUE}
levels(x)  # prints the levels, in order
nlevels(x) # check the number of levels
```

## Factors

- Can define the order of the factors
```{r, echo = TRUE}
x <- factor(c("low","high","medium","medium","high","low"), 
            levels = c("low","medium","high"))
levels(x)
```
- Can order the levels as numericals
```{r, echo = TRUE}
min(x) 
x <- factor(c("low","high","medium","medium","high","low"), 
            levels = c("low","medium","high"), ordered = TRUE)
min(x)
```

## Converting factors

- **Careful** whe converting factors to numeric
```{r, echo = TRUE}
x <- factor(c(1.2,3.5,5))
as.numeric(x)
```
<!-- ask why? has to do with the fact that the content of factors are stored as integers, regardless what they are-->
- Reason: factors are ultimately stored as integers
- Workaround
```{r, echo = TRUE}
as.numeric(levels(x))
```

## Usefulness of factors

<!-- To understand how useful factors are for data analysis, let's load one of our data set -->
```{r, include = FALSE}
setwd("/Users/hugolhuillier/Dropbox/Teaching/intro-prog/2017-2018/4-8-r")
```
- By default strings are stored as factor
```{r, echo = TRUE}
# DO NOT FORGET TO BE IN THE CORRECT DIRECTORY WITH setwd()
d <- read.csv(file = "data/sample.csv")
str(d)
```

## Usefulness of factors

- Factors are treated differently than numbers and strings by many functions
```{r, echo = TRUE}
summary(d) # for factors, counts the occurence of each level
```

## Factors: exercice #1 

- The function `table()` tabulates observation and can be used to create bar plots
- **Example**
```{r, echo = TRUE, fig.align='center'}
barplot(table(d$Group))
```
- **Exercice**: use `factor()` to modify the order and plot the control group last

## Factors: solution 

```{r, echo = TRUE, fig.align='center'}
# use the indexing [-1] to remove the first element of levels(d$Group), then combine via c() 
# identical to levels = c("Control", "Treatment1", "Treatment2")
d$Group <- factor(d$Group, levels = c(levels(d$Group)[-1], levels(d$Group)[1]))
barplot(table(d$Group))
```

## Working with factors 

- Clearly the gender entry has been coded with errors: 
```{r, echo = TRUE}
table(d$Gender)
```
- `F` and `M` should have been recorded as `f` and `m`
```{r, echo = TRUE}
levels(d$Gender)[2] <- "f"
levels(d$Gender)[3] <- "m" # careful, [3] not [4]! 
table(d$Gender)
```
