---
title: "Introduction to Programming"
subtitle: "<h3> *R: Data Analysis (data visualization)* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE}
require(tidyverse)
require(nycflights13)
require(gapminder)
require(Lahman)
```

# Packages 

## Tidyverse 

- A package: a collection of functions, data, and documentation that extends the capabilities of base R
```{r, eval = FALSE}
install.packages("tidyverse") # install a package 
library(tidyverse)            # load the library / package
```
- `tidyverse`: a collection of packages that work very well together

# Data visualization 

## `ggplot2`

- Seems complicated to use 
- Generates quickly amazingly detailed plots 
- Practial question: 

  > Do cars with big engines use more fuel than cars with small engines?
  
- Will answer via the `mpg` data frame (`displ`: size, `hwy`: car's fuel effiency)

```{r, echo = TRUE}
head(mpg)
```

<!-- 
hwy, a car’s fuel efficiency on the highway, in miles per gallon (mpg):  a car with a low fuel efficiency consumes more fuel than a car with a high fuel efficiency when they travel the same distance.
-->

## `ggplot2`: default template 

- Every `ggplot2` plot follows
```{r, echo = TRUE, eval = FALSE}
ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
- `ggplot()` creates a coordinate system that you can add layers to
- `data`: the dataset to be used in the graph 
- `GEOM_FUNCTION`: adds a layer to the plot
- `mapping` and `aes`: defines how variables in your dataset are mapped to visual properties

# Scatter plots

## Scatter plot: basics

- To plot `hwy` against `displ` 

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy))
```

<!-- negative relationship between engine size and fuel efficiency: cars with big engines use more fuel. -->

- Outliers? Could be hybrids?

## Scatter plot: adding another variable

- Add a third variable to a 2d scatterplot by mapping it to an _aesthetic_
- Aesthetic: visual property of the objects in the plot 
- To map an aesthetic to a variable, associate the name of the aesthetic to the name of the variable inside `aes()`

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

- Can try with `size`, `alpha` or `shape`

## Scatter plot: general aesthetic properties

- Can also set the aesthetic properties of your geom manually 
```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

- Can specify

    1. color (associated to a string)
    1. size (associated to a size in mm)
    1. shape (associated to a number between 0 and 20)
    
## Scatter plot: exercice 

1. In `mpg`, identify which variables are categorical and which are continuous 
2. Map a continuous variable to color inside `aes` 

## Scatter plot: exercice 

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = cty))
```

## Scatter plot: exercice 

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = year))
```

# Facets & grids

## Facets 

- Can also split the plot into facets (=subplots), via `facet_wrap()`
- First argument: formula, created with `~` followed by a variable name -- necessarily discrete (but not necessarily categorical)!
```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
``` 

## Grids

- Before: split with respect to one variable 
- Can also split with two variables with `facet_grid()`
```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```    
<!-- cyl: cilinders, drv: cars' drivetrain 
4 stands for four-wheel drive, f for front-wheel drive, and r for rear-wheel drive.-->

# Geometerical objects

## Geometerical objects

- `ggplot2` is only about adding layers to a plot
- A _geom_: geometrical object used by `ggplot2` to represent data 
- Ex: plot `hwy` against `displ`, this time using `geom_smooth` instead of `geom_point`

## Geometrical objects 

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy))
```   

## Geometerical objects 

- `ggplot2` is only about adding layers to a plot
- A _geom_: geometrical object used by `ggplot2` to represent data 
- Some mapping arguments work only for some geom (tried to shape a line?)
- In total, about 30 geoms provided by `ggplot2`

## Geometrical objects 

- For some geoms (e.g. smooth), can you the _group_ aesthetic
- Group: `ggplot2` will group the data for these geoms whenever used with a discrete variable, without adding a legend or distinguishing features

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```   

## Multiple geoms 

- Of course, can add multiple layers to the same plot

```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```  

## Multiple geoms 

- In the previous command, some duplication of our code 
```{r, echo = TRUE, eval = FALSE} 
geom_point(mapping = aes(x = displ, y = hwy)) + geom_smooth(mapping = aes(x = displ, y = hwy))
```
- Instead, can define a global mapping directly in `ggplot()`
```{r, fig.align='center', echo = TRUE, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + geom_point() + geom_smooth()
```  

## Multiple geoms 

- If place mappings in a geom function in addition to a global mapping, `ggplot2` will extend or overwrite the global mappings for that layer only
```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

## Multiple geoms 

- Can also specify data for each layer 
```{r, fig.align='center', echo = TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## Geoms: exercices 

```{r, include = FALSE}
require(gridExtra)
```

- Reproduce these plots
```{r, fig.align='center', echo = FALSE}
p1 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
        geom_point() + 
        geom_smooth(se = FALSE)
p2 <- ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
        geom_point(mapping = aes(color = drv)) + 
        geom_smooth(mapping = aes(linetype = drv), se = FALSE)
grid.arrange(p1,p2)
```

## Geoms: solutions

```{r, eval = FALSE}
# first plot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
# second plot 
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = drv)) + 
  geom_smooth(mapping = aes(linetype = drv), se = FALSE)
```

# Statistical transformations 

## Bar charts 

- Based on `diamonds` dataset, with variables s.a. `price`, `carat`, `color`, `clarity`, `cut`...
- Bar chart with `geom_bar`
- Note: only one variable in the mapping
```{r, fig.align='center', echo = TRUE}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut))
```

## Bar charts

- On the y-axis: _counts_ the number of diamonds / category 
- "Count" is the default `stat` associated to `geom_bar`

```{r, fig.align='center', echo = TRUE}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

- `ggplot2` provides over 20 stats (e.g. `stat_smooth` is the stat associated to `geom_smooth`)

## Bar charts: exercice 

- Run the cell below and understand what `geom_col` do?
```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + geom_col(mapping = aes(x = cut, y = carat))
```
- Run the cell below and undestand why we need `group` 
```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, y = ..prop.., group = "x"))
```

<!-- 
The default for geom_bar is to group by the x variable in order to separately count the number of rows in each level of the x variable. 
If we want proportions, then we need to consider all levels of cut together. This is not the default behavior; the data are first grouped by x, so each level of x is considered separately. The proportion of x1 in x1 is 100%, as is the proportion of x2 in x2 etc.
Group="whatever" overrides the default behavior, which is to group by the x variable. 
-->

# Position adjustments 

## Bar charts (cont.)

- As for scatterplot, can add another variable to a bar chart 
```{r, echo = TRUE, eval = FALSE}
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = cut))
ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = clarity))
```
```{r, fig.align='center', echo = FALSE}
p1 <- ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = cut))
p2 <- ggplot(data = diamonds) + geom_bar(mapping = aes(x = cut, fill = clarity))
grid.arrange(p1,p2)
```

## Bar charts & position adjustment 

- By default, the bars are stacked 
- The stacking is performed by the position adjustment argument, specified by `position`

    1. `position = "identity"`: overlaps the bars
    1. `position = "fill"`: works like stacking, but makes each bars the same heigh
    1. `position = "dodge"`: places overlapping objects beside one another
    
## Bar charts & position adjustment 

```{r, fig.align='center', echo = FALSE}
g  <- ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity))
p1 <-  g + geom_bar(alpha = 1/5, position = "identity")
p2 <- ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(fill = NA, position = "identity")
p3 <- g + geom_bar(position = "fill")
p4 <- g + geom_bar(position = "dodge")
grid.arrange(p1,p2,p3,p4,ncol=2)
```
```{r, fig.align='center', eval = FALSE, echo = TRUE}
g  <- ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity))
g + geom_bar(alpha = 1/5, position = "identity")
g + geom_bar(fill = NA, position = "identity")
g + geom_bar(position = "fill")
g + geom_bar(position = "dodge")
```

## Position adjustments: exercice 

- Using `geom_boxplot`, play with the position argument. Specifically, based on the `mpg` dataframe, do a boxplot of `hwy` on `class`, differenting by `drv`. 

    1. What is the default position?
    1. Which of `fill`, `dodge` and `ìdentitiy` work?
    
## Position adjustments: solution 
    
```{r, fig.align='center', echo = FALSE}
g <- ggplot(data = mpg, mapping = aes(x = class, y = hwy, fill = drv))
p1 <- g + geom_boxplot()
p2 <- g + geom_boxplot(position = "identity")
grid.arrange(p1,p2)
```
```{r, fig.align='center', eval = FALSE, echo = TRUE}
g <- ggplot(data = mpg, mapping = aes(x = class, y = hwy, fill = drv))
g + geom_boxplot()
g + geom_boxplot(position = "identity")
```

# Coordinate systems 

## Coordinate systems

- Default coordinate system is the Cartesian coordinate system 
- Others: 

    - `coord_flip()`: switches the x and y axes 
    - `coord_polat()`: switches to polar coordinates
    
## Coordinate systems 

```{r, fig.align='center', echo = TRUE}
# switches x and y
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + geom_boxplot() + coord_flip()
```

## Coordinate systems 

```{r, fig.align='center', echo = TRUE}
# polar coordinates
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut), show.legend = FALSE) + 
  labs(x = NULL, y = NULL) + 
  coord_polar()
```

