---
title: "Introduction to Programming"
subtitle: "<h3> *R: Input / Output* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Reading `.csv`

<!-- how to read data from a .csv and write to a new .csv, and explore the arguments that allow you read and write the data correctly for your needs --> 

## Loading the data

- `read.csv` loads `.csv` file as data frame
```{r, echo = TRUE}
setwd("/Users/hugolhuillier/Dropbox/Teaching/intro-prog/2017-2018/4-8-r/")
d <- read.csv(file = "data/car-speeds.csv")     # data on car speeds and color
str(d)
```
- By default, the default delimiter in `.csv` files is a comma; can specify an arbitrary delimiter via the `sep` argument in `read.csv()`

## Useful arguments

- `header`: by default, `TRUE` $\Rightarrow$ the first row of values in the `.csv` file is set as column names
- `stringsAsFactors`: by default, `TRUE`
<!-- it could be that the strings are not categorical, e.g. names. in this case, better to use FALSE since there are functions that works only on strings and not factors-->
- `as.is`: extension of `stringsAsFactors` that gives control over individual columns
```{r, echo = TRUE}
d <- read.csv(file="data/car-speeds.csv", as.is = 1)
str(d)
```
<!-- the first column was importer AS IT IS, while the third one ws converted into factors-->
- `strip.white`: if the data was mis-recorded with spaces, `R` removes these 

## `strip.white`: example

- In our dataset, seems that `Red` has been mis-recorded sometimes 
```{r, echo = TRUE}
unique(d$Color)       # remember that the Color column is stored as string => cannot use levels()
d <- read.csv(file="data/car-speeds.csv", as.is = 1, strip.white = TRUE, sep = ",")
unique(d$Color) 
```

# Writing `.csv` file

## Writing the data 

- Data frame are saved in `.csv` via `write.csv`; two arguments 

    1. the data frame to be saved 
    1. the name of the file 
    
```{r, echo = TRUE}
write.csv(d, file = "data/car-speeds-cleaned.csv")
```

## Useful arguments 
 
- `row.names`: by default `TRUE`, the first column of the new `.csv` file will contain numbers (= # of the rows)
- `na`: remember that `NA` is `R`-specific; can specify with which value to replace `NA` with the argument `na`

