---
title: "Introduction to Programming"
subtitle: "<h3> *R: Using logicals* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Conditional statement

- Logicals (or booleans) are useful to adapt your program to different scenari
```{r, echo = TRUE}
x <- 0
if (x > 1) {
  print("x greater than 1")
} else { 
  print("x smaller than 1")
}
```
- _Conditional statement_: a statement that might or might not be executed depending on whether a test is true or false
<!-- 
if the condition under the bracket (x > 1) is true, then execute the first block
else, i.e. if it is false, excute the second block
-->

## Conditional statements

- Do not have to include an `else` statement
```{r, echo = TRUE}
if (x > 1) {
  print("x greater than 1")
} 
```
- Can also add multiple condition 
```{r, echo = TRUE}
if (x > 1) {
  print("x greater than 1")
} else if (x == 1) {
  print("x equal 1")
} else { 
  print("x smaller than 1")
}
```

## Comparisons 

- Multiple operators to make logical comparison in `R`

    - `x == y`: returns `TRUE` if $x = y$
    - `x != y`: returns `TRUE` if $x \neq y$
    - `x > y`: returns `TRUE` if $x > y$ (id. for `<`)
    - `x >= y`: returns `TRUE` if $x \geq y$ (id. for `<=`)
 
- `!` returns the opposite logical:c `!TRUE` returns `FALSE`

## Condition statement: exercice #1

- **Exercice**: write a function `sign` that returns the sign of a numeric 

## Condition statement: solution

- **Exercice**: write a function `sign` that returns the sign of a numeric (positive, negative, or zero)
```{r, echo = TRUE}
sign <- function(x) {
  if (x > 0) { 
    return(1)
  } else if (x == 0) { 
    return (0)
  } else { 
    return(-1)
  }
}
```

## Condition statement: exercice #2

- **Exercice**: write a function `plot_dist` that plots a boxplot (via `boxplot`) if the length of the vector to be plotted is greater than some threshold, and a stripchart (via `stripchart`) otherwise

## Condition statement: solution

- **Exercice**: write a function `plot_dist` that plots a boxplot (via `boxplot`) if the length of the vector to be plotted is greater than some threshold, and a stripchart (via `stripchart`) otherwise
```{r, echo = TRUE}
plot_dist <- function(x, threshold) { 
  if (length(x) > threshold) { 
    boxplot(x)
  } else { 
    stripchart(x)
  }
}
```
<!-- NOTE: if do not include return statement, R will automatically return the last instruction-->

## Condition statement: solution

- **Exercice**: write a function `plot_dist` that plots a boxplot (via `boxplot`) if the length of the vector to be plotted is greater than some threshold, and a stripchart (via `stripchart`) otherwise
```{r, fig.align="center",echo = TRUE}
x <- rnorm(15)
par(mfrow=c(1,2))
plot_dist(x, 20)
plot_dist(x, 10)
```
```{r, include = FALSE}
par(mfrow=c(1,1))
```

## Multiple conditions 

- Can combine conditions, as in maths with the $\cup$ and $\cap$ operators

    - `&` stands for `and` (as $\cap$)
    - `|` stands for `or` (as $\cup$)
    
<!-- as in maths, 'OR' means“either or both” -->
```{r, echo = TRUE}
x <- 1
if (x > 0 & x < 0) {
  print("this will never print")
} 
if (x >= 0 | x <= 0) { 
  print("this will always print")
}
```

## Multiple conditions: exercice #3 

- **Exercice**: edit the `plot_dist` function to add an additional argument `use_boxplot`; when set to true (default behavior), a large data set will be plotted as a boxplot; otherwise, it will be plotted as a histogram (using `hist()`)

## Multiple conditions: solution

- **Exercice**: edit the `plot_dist` function to add an additional argument `use_boxplot`; when set to true, a large data set will be plotted as a boxplot; otherwise, it will be plotted as a histogram (using `hist()`)
```{r, echo = TRUE}
plot_dist <- function(x, threshold, use_boxplot = TRUE) {
   if (length(x) > threshold & use_boxplot) {
       boxplot(x)
   } else if (length(x) > threshold & !use_boxplot) {
       hist(x)
   } else {
       stripchart(x)
   }
}
```

## Multiple conditions: solution

```{r, fig.align="center",echo = TRUE}
x <- rnorm(15)
par(mfrow=c(1,3))
plot_dist(x, 20)
plot_dist(x, 10)
plot_dist(x, 10, use_boxplot = FALSE)
```
```{r, include = FALSE}
par(mfrow=c(1,1))
```

