---
title: "Introduction to Programming"
subtitle: "<h3> *R: Loops* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The BLRs: again & again 

- Loops are one of the easiest way to automate your work
- Allows you to repeat some instructions without re-writing the them every time

# `For` loops

## Automate your work

- `For` loops lets you iterate over different items and apply some instructions to each of these items
```{r, echo = TRUE}
# DO NOT TRY THIS AT HOME
words <- c("These", "are", "different", "words")
print(words[1])
print(words[2])
print(words[3])
print(words[4])
```

>- **Exercice:** write the exact same code for a vector containing 100 entries
    
## Automate your work

- `For` loops lets you iterate over different items and apply some instructions to each of these items
```{r, echo = TRUE}
# DO NOT TRY THIS AT HOME
words <- c("These", "are", "different", "words")
print(words[1])
print(words[2])
print(words[3])
print(words[4])
```
- Terrible function 
    
    >1. Doesn't scale
    >2. Requires too much manpower!

<!-- if we give it a longer vector, it only prints part of the data, and if we give it a shorter input, it returns NA --> 

## Automate your work

- Whenever you want to do something repetitive: **use a loop**
```{r, echo = TRUE}
words <- c("Now", "it", "works", "regardless", "of", "the", "length", "of", "the", "vector")
for (word in words) {
    print(word)
}
```

## Automate your work

- Alternatively
```{r, echo = TRUE}
words <- c("Now", "it", "works", "regardless", "of", "the", "length", "of", "the", "vector")
for (i in 1:length(words)) {
    print(words[i])
}
```

## `For` loops: general structure

- A `for` loop always write 
```{r, eval = FALSE}
for (variable in collection) {
  do things with variable
}
```

## `For` loops: exercice 

- Write a functiont that takes a numerical vector as argument, and sum its elements via a `for` loop
- Test your function using the `R` function `sum`

## `For` loops: solution 

- Write a functiont that takes a numerical vector as argument, and sum its elements via a `for` loop
```{r, echo = TRUE}
require(testthat)
own_sum <- function(vec) {
  s   <- 0 
  for (v in vec) { 
    s <- s + v  
  }
  return(s)
}
```
```{r, echo = TRUE}
x <- c(1, 2, 3)
test_that("Testing own_sum function", {
  expect_true( own_sum(x) == sum(x), "own_sum == sum")
})  
```
<!-- DO NOT FORGET TO COMMENT THE FUNCTION 
1. start with s = 0 
1. then for each iteration, v is assigned a new element in the vector (in order)
1. then we set s equal to its curent value + the value of v 

RECALL: once again, should always use `R` function over our own function if they have the same purpose, because R functions are much faster!
Reason: sum is much faster than the interpreted for loop because sum is coded in C to work with a vector of numbers. Many of R’s functions work this way; the loop is hidden from you in C.
-->

## `For` loops

- If you are not sure what the sum is doing, you can print the elements in the collection via the function `print`
```{r, echo = TRUE}
for (v in c(1, 2, 3)){
  print(v)
}
```
- Note: `v` is a (loop) variable, used to record progress in a loop; the variable exists after the loop is over!
```{r, echo = TRUE}
v
```

# Other loops 

## `while` loop

- A logical condition determines whether the iteration should go on

    1. if the logical condition = `TRUE`: the instructions are executed 
    1. otherwise, stop

```{r, eval = FALSE}
while(logical condition){
  do things
}
```

## `while` loop: an example

- **Ex**
```{r, echo = TRUE}
x <- 1 
while (x < 5) { 
  x <- x + 1
  print(x)
}
```

- Careful: can easily end up with infinite sum (e.g. if forgot to update with `x <- x + 1`)

## Loops: `next` and `break`

- Can stop the iteration with `break`
```{r, echo = TRUE}
x <- 1 
while (x < 5) {
  x <- x + 1 
  if (x == 3) { 
    break       # stop and exit the loop
  }
  print(x)
}
```
- Can skip steps with `next` 
```{r, echo = TRUE}
x <- 1 
while (x < 5) {
  x <- x + 1 
  if (x == 3) { 
    next       # any instructions after next are not executed for this iteration
  }
  print(x)
}
```

## `repeat` loop

- Very similar to `while` loops, but the block of instructions is executed at least one, regardless whether the logical condition is met, and the condition is hard-coded 
```{r, echo = TRUE} 
x <- 1 
repeat {
  x <- x + 1 
  print(x)
  if (x == 5) {
    break
  }
}
```

## `repeat` loop: exercice 

- Reproduce the loop that printed `2, 4, 5` with a `repeat` loop

## `repeat` loop: solution 

- Reproduce the loop that printed `2, 4, 5` with a `repeat` loop
```{r, echo = TRUE} 
x <- 1 
repeat {
  x <- x + 1 
  if (x == 3) { 
    next
  }
  print(x)
  if (x == 5) {
    break
  }
}
```


# Vectorized operations

## Sum vs. vectors (= intro to profiling)

- `R` hates raw sum, but is very good at handling vectors 
- **Ex**: could write our own function to add two vectors: 

$$
z = x+y \quad \text{for} \, x,y \in V.
$$
```{r, echo = TRUE}
vec_add <- function(a, b){
  res <- numeric(length = length(a))  # create a new vector to store the sum
  for (i in seq_along(a)) {           # seq_along generates indices for each element in `a` 
                                      # equivalent to 1:(length(a))
    res[i] <- a[i] + b[i]             # for each entry in the vector, store the sum in res
  }
  return(res)
}
a   <- 1:10e5; b <- 1:10e5
res <- vec_add(a, b)
```

## Sum vs. vectors (= intro to profiling)

- `vec_add` is a terrible function: 

    1. take only two arguments, so cannot do $x + y + z$ for $x,y,z \in V$
    1. will not throw an error if a and b are not of the same length

- `R` already has a vectorized functions to do that: `+`
```{r, echo = TRUE}
res2 <- a + b
```
```{r, echo = TRUE}
test_that("Sum of vectors", {
  expect_true( all(res == res2), "res == res2")
})  
```

## Timing your code

- Time the execution of your code via `system.time()`
```{r, echo = TRUE}
system.time(result <- vec_add(a, b))
system.time(result <- a + b)
```

## Vector recycling

- If perform an operation on vectors of unequal length, `R` will "recycle" elements of the shorter vector(s) to match the longest one 
```{r, echo = TRUE}
a <- 1:10; b <- 1:5 
a + b
```
<!-- 
When R reaches the end of the shorter vector b, it starts again at the first element of b and continues until it reaches the last element of the longest vector a
--> 
- Recycling occurs silently if the lengths of the two vectors are multiple 
```{r, echo = TRUE}
a <- 1:10; b <- 1:3
a + b
```
- Why? Very useful when want to perform the same operation on every element of a vector
```{r, echo = TRUE}
a <- 1:10; b <- 5 
a * b
```
<!-- Remember there are no scalars in R, so b is actually a vector of length 1; in order to add its value to every element of a, it is recycled to match the length of a. -->

## `for` or `apply` ?

- `For` loop: same function calls to a collection of objects
- Very similar to `apply`: each of the following functions take an argument `FUN`, a function to be applied to each element of the object 

    - `apply`: apply the function over the margins of an array (rows or columns of a matrix)
    - `lapply`: apply the function over an object an return list
    - `sapply`: apply the function over an object and retur a simplified object (an array) if possible
    - `vapply`: similar to `sapply` but can specify the type of object to be returned

<!-- Deciding whether to use for or one of the apply family is really personal preference;
1. for loops are often more natural 
1. for several related operations, a for loop will avoid you having to pass in a lot of extra arguments to your function 
NOTE: in most cases, apply functins are not much faster than for loops -->

## Rules for `R` loops 

1. Do not use loop when a vectorized object exists <!-- already gave an example earlier on -->
1. Do not grow objects during the loop
1. Instead, allocate an object to hold the results and fill it in during the loop

## Rules for `R` loops: example

```{r, echo = TRUE}
col_median_slow <- function(mat){ 
  # function that computes the median for each column in a matrix 
  for (i in 1:(dim(x)[2])){ 
      if (i == 1) { 
        out <- median(x[,i])
      } else {
        out <- cbind(out, median(x[,i]))
      }
  }
  return(out)
}
```    
```{r, echo = TRUE}
col_median_fast <- function(mat){ 
  # function that computes the median for each column in a matrix 
  out <- numeric(length = dim(mat)[2])
  for (i in 1:(dim(x)[2])){ 
    out[i] <- median(x[,i])
  }
  return(out)
}
```  
<!-- in the slow version, add a new column to out at each iteration, hence at each iteration, R has to create a new object and copy across the information just to add a new element or row/column 
-->
<!-- FINAL NOTE 
  note how often we use function. in the two examples above, could have written these examples w/out functions, but always better to do it in a function (more re-usable + could time etc.)
-->

## Rules for `R` loops: example

```{r, echo = TRUE}
x <- matrix(1:10e5, ncol = 10e3)
system.time(result <- col_median_slow(x))
system.time(result <- col_median_fast(x))
```
