---
title: "Introduction to Programming"
subtitle: "<h3> *R: Introduction* </h3>"
author: Hugo Lhuillier
date: Master in Economics, Sciences Po
output: 
  revealjs::revealjs_presentation:
    center: true 
    highlight: pygments
    css: my-style.css
    transition: slide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R: where the fun begins

Why are we using `R`? 

- Better than Stata by ANY metric
- Free 
- Extremely popular amongst scientists, in particular statistians and economists
- Exists a large library of external packages

# Variables
## A very gentle intro. to variables

* A variable: a container with a name
* To create a variable 

```{r, echo = TRUE}
weight_kg <- 55
weight_kg
```

- Can treat the variable like a regular number

```{r, echo = TRUE}
weight_kg + 1
```

- Can change an variable's value by assigning it a new value

```{r, echo = TRUE}
# changing the variable weight_kg from 55 to 60
weight_kg <- 60
```

<!-- Once a variable is created, we can use the variable name to refer to the value it was assigned. The variable name now acts as a tag. Whenever R reads that tag (weight_kg), it substitutes the value (55). 
In general, R will print to the console any object returned by a function or operation unless we assign it to a variable.-->

## A very gentle intro. to variables

- R only stores the value, not the calculation used to create a variable ($\neq$ Excel)
```{r, echo = TRUE}
weight_lb <- 2.2 * weight_kg
c(weight_kg, weight_lb)
weight_kg <- 80 
c(weight_kg, weight_lb)
```

<!-- 
1. `c` is also a function (probably the most used function in R), stands for combine
2. 1:5 = c(1, 2, 3, 4, 5) but much faster
-->

## Variables' names: conventions

- Some conventions on the name of variables 
    1. start with lower case letters 
    1. separate words with underscores
    1. use only lowercase letters, underscores, and numbers

# Motivating example: analyzing data 

## The data

- In the `data` directory, several panel data: data on inflammation in patients who have been given a new treatment for several days
- Goal: load the data, calculate the average value of inflammation per day, plot the results

## Loading data 

- Need to move to the correct directory, then load the data
```{r, echo = TRUE}
setwd("/Users/hugolhuillier/Dropbox/Teaching/intro-prog/2017-2018/4-8-r/data/")
d = read.csv(file = "inflammation-01.csv", header = FALSE)
```
- `header` is an optional argument, with a default value = `TRUE`. Get the list of all the optional arguments by running `help(read.csv)`
<!-- 
both setwd() and read.csv() are functions that takes some arguments 
1. the first argument of both functions is a String => put quotes 
2. the second argument of read.csv is what we call a Boolean value (either true or false). Header:  whether the first line of the file contains names for the columns of data

d = data frame. more on this later: but basically, like an excel sheet.

NOTE: if forgot to add `d` before the `read.csv`, `R` does not know what to do with the function's output. will display the full contents of the file inflammation-01.csv in the console. instead here, `d` is a variable containing the data that was in `inflammation-01`
-->
- Display the first lines of the data set with `head` 
```{r, echo = TRUE}
head(d, n = 3L)
```

# Manipulating data
## Indexing
- To take a subset of the data set, provide an **index** in square bracket: `[# row, # column]` 
```{r, echo = TRUE} 
d[1,1]                      # first row, first column
d[c(1, 3, 5), c(10, 20)]    # rows (1, 3 and 5), columns (10 and 20)
d[1, 1:5]                   # columns from (1 to 5) and row 1
d[, 1]                      # all columns from row 1
```

## Indexing: exercice

- In our data set, each row is a patient, each column is a day, such that `d[1,1]` is the inflammation measured on patient 1 on day 1
- **Exercice**: given that `min(data)`, `max(data)`, `mean(data)` are functions returning the equivalent statistics on d[c(2, 4, 10),]`data`, finds 
    1. the minimum inflammation on day 1 across all patients
    1. the maximum inflammation experienced by patient 5 (across all days)
    1. the maximum inflammation on days 4, 8 and 12 across all patients
    1. the minimum inflammation experienced by patients 3 and 6 from day 1 to 5
    1. the mean inflammation experienced by patients 2, 4 and 10 (across all days)

## Indexing: solution

```{r, echo = TRUE}
min(d[, 1])               # the min inflammation on day
max(d[5, ])               # the max inflammation for patient 5 
max(d[, c(4, 8, 12)])     # the max inflammation on days 4, 8 and 12 
min(d[c(3, 6), 1:5])      #  the min inflammation for patients 3 and 6 from day 1 to 5
rowMeans(d[c(2, 4, 10),]) # the mean inflammation for patients 2, 4 and 10
```

## Sufficient statistics 

- Faster way to get some sufficient statistics (**by columns**): `summary`
```{r, echo = TRUE}
summary(d[, 1:5])
```
<!-- ask what would the command 
summary(d[1:4,]) 
do for instance?-->

## The `apply` function

- What if we want some info, say the median, for each partient (= row)? No such things as `rowMedian`
- `apply`: repeat a function on all of the rows (`MARGIN = 1`) or columns (`MARGIN = 2`) of a data frame
```{r, echo = TRUE}
apply(d, 1, median)
```

## The `apply` function: exercice

- **Exercice**: compute in two different ways the mean for the first 10 patients of our data 

## The `apply` function: solution

- **Exercice**: compute in two different ways the mean for the first 10 patients of our data 
```{r, echo = TRUE}
apply(d[1:10,], 1, mean)
rowMeans(d[1:10,])
```
- __NOTE__: whenever a function is coded in base R, use it!

# $\heartsuit$ plotting $\heartsuit$
## $\heartsuit$ plotting $\heartsuit$

- Quote from a wise man

> what the point of doing economics is if plotting your data you cannot, hmm?  

## $\heartsuit$ plotting $\heartsuit$

- `R`'s plot are very nice! 
```{r, fig.align='center', echo = TRUE}
plot(apply(d, 2, max), xlab = "day", ylab = "maximum", main = "maximum inflammation by day")
```

## $\heartsuit$ plotting $\heartsuit$

- `R`'s plot are **very** nice! 

```{r, fig.align='center', echo = FALSE}
boxplot(d, main = "Summary")
```